\documentclass{letter}
\signature{Co-authors}
\begin{document}
\begin{letter}{}
\opening{Dear Reviewers:}

Thank you for your constructive comments. We have addressed the expressed concerns as follows:
\begin{enumerate}
\item Page 1: "How did you select PCA and CX as the two low rank factorization to
  investigate?" --- We address this in the introduction: the PCA is useful for
  capturing variance, and the CX is desirable for its interpretability.
\item Page 2: "What is RLA?" --- We now explicitly state randomized linear algebra.
\item Page 2: "The description of Algorithm 1 requires the reader to understand what is
  meant by several terms, some of which are not explained and others explained
  later in the paper. It would have been helpful to have a description similar
  to algorithm 2." --- We have expanded upon the explanation of Algorithm 1 to provide an idea of why it works.
\item Page 3: "What does `close to the metal C' refer to?" --- We meant highly optimized, but we have removed this term.
\item Page 3: "Why did you select Spark?" --- Spark is a popular distributed data analytics framework, and we are interested in quantifying the gap between it and traditional HPC codes when applied to linear algebra. We clarify this in the introduction.
\item Page 3: "`Non-zeros' - are the matrices sparse?" --- We now state that the MSI matrix is sparse and the climate matrix is dense.
\item Page 3: "`1TB data set' - but you have not yet described the data." --- We describe the data when it is introduced now.
\item Page 4: "Near top, left column - what is the QR step?" --- We specify it is Step 7 in Algorithm 3
\item Page 5: "What are the matrices used in the MSI and climate data to which you are applying CX and PCA?" --- We discuss how the matrices are constructed from the respective datasets when we introduce them.
\item Page 8: "Right column, analysis of MSI data - there are three numbers, but only two materials - CH2 and CH2O. what happened to the matereial with the highest score around 380?" --- We only have knowledge of the materials we discussed.
\item Page 9: "Who evaluated the analysis results? Are there biologists and climate scientists involved in this effort?" --- Several of the co-authors are biologists, and several are climate scientists. They provided interpretations of the respective decompositions.
\end{enumerate}
Additionally, we updated the figure comparing the runtimes of the PCA algorithms on the XC40 and EC2.
\closing{Sincerely,}
\end{letter}
\end{document}

